---
always_allow_html: true
output:
  rmarkdown::github_document: default
  always_allow_html: default
  output:
  md_document:
    variant: gfm
knit: (function(inputFile, encoding) {
  rmarkdown::render(inputFile, encoding = encoding, output_dir = "../") })
date: "`r Sys.Date()`"
title: "_0_ccmhc_data_review_from_rfiles"
---


```{r setup, include=FALSE}
base_dir <- "../" # where output should default to
base_url <- "" # keep as is
fig_path <- "_plot_images/" # where image output should go

knitr::opts_knit$set(base.dir = base_dir)
knitr::opts_chunk$set(fig.path = fig_path,
                      cache.path = 'cache/',  #log files go here
                      message=FALSE, warning=FALSE, cache = FALSE) 
```  

# 1. Setup & Data Ingestion
## 1.1 Libraries

```{r}
library(dplyr)       # Data manipulation
library(stringr)     # String manipulation
library(tidyr)       # Data pivoting
library(readr)       # Reading data
library(kableExtra)  # HTML Table formatting
library(ggplot2)     # Visualizations
library(ggpubr)      # Publication-ready plots
library(rstatix)     # T-tests and effect sizes
library(broom)       # Tidy statistical outputs
library(forcats)     # Factor manipulation
library(effectsize)  # effect size & omega squared

```

## 1.2 Load Data
### 1.2.0 !! Updated Data File:
- **UPDATED DATA FILE** - SEFF_Updated...
  - Both Surveys are connected to a single participant ID (Pre and Post) and collated in one spreadsheet
  - Will separate each into its own table, for clarity/checks and file write out. 
  - Rejoin and save data files for analysis scripts.

- **Structure**
  - Study_ID (Joined ID across pre and post, unique to participant)
  - Group Assignment (Control or Intervention)
  - Stress Scale: **PSS** - *Perceived Stress Scale* - 10 Questions
  - SEFF Scale: **EFF** - *Eearly Intervention Parent Self Efficacy* -  14 Questions
  - Columns as Pre --> Post 

# NEW FILES FROM R CHECK
```{r}
## raw_data in CSV format... means we get the numbers (and also some added data - rather than LABELS (strongly agree etc...))
# read RAWS
control.csv <- readr::read_csv("../_data/CONTROLPostMeasuresA_DATA_2026-08-24_1126.csv", col_names = TRUE)
pre.csv <- readr::read_csv("../_data/FamilyInfoAndPreMeas_DATA_2026-08-24_1122.csv", col_names = TRUE)
post.csv <- readr::read_csv("../_data/PostMeasuresAndSatis_DATA_2026-08-24_1123.csv", col_names = TRUE)
```


```{r}
control.csv %>% 
  kable(caption="Raw Data: Control Group") %>% 
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = F) %>%
  scroll_box(height = "400px", width = "100%")
```


```{r}
pre.csv %>% 
  kable(caption="Raw Data: Intervention - PRETEST") %>% 
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = F) %>%
  scroll_box(height = "400px", width = "100%")
```


```{r}
pre.PSS.sel <- pre.csv %>% 
  select("StudyID" = record_id,
            #"Group_Assignment" = "Intervention",
            contains("promis"))

pre.PSS.sel %>% 
  kable(caption="Raw Data: Intervention - PRETEST") %>% 
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = F) %>%
  scroll_box(height = "400px", width = "100%")
```
## Big Question: Are these numbers already reversed? 
 - And because we were exporting LABELS (Agree, Strongly Disagree) etc... did that NOT, or ALREADY do the reversing....???

```{r}
pre.PSS.mut <- pre.PSS.sel %>% 
  mutate(
    # CALC WITHOUT REVERSE 
    "PSS_RAW.sum" = rowSums(select(., 2:11), na.rm=TRUE),
    "PSS_RAW.mean" = rowMeans(select(.,2:11), na.rm=TRUE))

pre.PSS.REV.mut <- pre.PSS.sel %>%
  mutate(
    across(ends_with("_r"), ~ 5 - .)) %>% 
  mutate(
    "PSS_REV.sum" = rowSums(select(.,2:11), na.rm=TRUE),
    "PSS_REV.mean" = rowMeans(select(.,2:11), na.rm=TRUE)
    )

together <- cbind(
  pre.PSS.mut %>% select(StudyID, contains(c(".sum",".mean"))),
  pre.PSS.REV.mut %>% select(contains(c(".sum",".mean")))
  )

together %>% 
  kable(caption="BOTH METHODS") %>% 
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = F) %>%
  scroll_box(height = "400px", width = "100%")
```


```{r}
together.long <- 
together %>% 
  pivot_longer(-1, names_to = "Data_Method", values_to = "Value") %>% 
  mutate("Method" = ifelse(str_detect(Data_Method, "REV"), "Reverse", "Raw"),
"Calculation" = ifelse(str_detect(Data_Method, "sum"), "sum", "mean"))
```


```{r}
together.long %>% 
  ggplot(aes(x=Method, y = Value, color = Method, fill = Method))+
  geom_boxplot(alpha = .25)+
  geom_jitter()+
  facet_wrap(~Calculation, scales = "free")


```
### CTA THOUGHT:
- This looks like I should be using the **RAW** column output; (if going from RAW data) e.g.
- **RED CAP ALREADY REVERSES THE QUESTIONS** (They are tagged as reverse coded)
- Comparing this to the other data BELOW::

## From previous data wrangling... 
- StudyID is REDCAP ID!!!
- Not 100% I trust that

```{r}
analytic.files <- readRDS("../_output/analytic_files_scored_surveys.Rds")
PSS.scored <- analytic.files$PSS.Scored

check_joined <- left_join(
  x = PSS.scored %>% select(StudyID, "PRE_Stress.sum" = PRE_Stress.Total, "PRE_Stress.mean" = PRE_Stress.Mean),
  y = together,
  by = "StudyID"
)

check_joined.m <- check_joined %>% pivot_longer(-1, names_to = "Data_Method", values_to = "Value") %>% 
  mutate("Method" = ifelse(str_detect(Data_Method, "PRE"), "original", "new"),
         "REVERSE" = ifelse(str_detect(Data_Method, "REV"), "Reverse", "Raw"),
         "Calculation" = ifelse(str_detect(Data_Method, "sum"), "sum", "mean"))

check_joined.m %>% 
  filter(!is.na(Value)) %>% 
  kable(caption="Comparing Output from REDCAP") %>% 
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = F) %>%
  scroll_box(height = "400px", width = "100%")

```


```{r}
check_joined.m %>% 
  ggplot(aes(x=Method, y = Value, color = REVERSE, fill = REVERSE))+
  geom_boxplot(alpha = .25)+
  geom_jitter()+
  facet_wrap(~Calculation, scales = "free")
  
```

### -------

## Other Data from CSV

```{r}
pre.EFF.sel <- pre.csv %>% 
  select("StudyID" = record_id,
            #"Group_Assignment" = "Intervention",
            sickchild:pre_parent_self_efficacy_complete) %>% 
  rename_with(~paste0("EFF_Q", 1:16, "_", .), .cols = 2:17)

pre.EFF.sel %>% kable(caption="PRE SURVEY - RAW Output from REDCAP") %>% 
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = F) %>%
  scroll_box(height = "400px", width = "100%")
```


```{r}
post.csv %>% 
  kable(caption="POST SURVEY - RAW Output from REDCAP") %>% 
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = F) %>%
  scroll_box(height = "400px", width = "100%")
```

## ALL DATA FILE 

```{r}
# Read in raw dataset
ALL_DATA.raw <- read_csv("../_data/stress and SE ALL blind.csv", col_names = TRUE)

string_stub <- function(x){paste0(str_sub(x,1,17), "...", str_sub(x,-2,-1))}
#string_stub("In the past month... How often have you felt that you were unable to control the important things in your life?...6")

ALL_DATA.raw %>% 
  rename_with( ~ifelse(str_length(.) > 30, yes = string_stub(.), .)) %>% 
  kable(caption="All Survey Data - Linked to Study ID (Anon) - Shortened Column Names") %>% 
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = F) %>%
  scroll_box(height = "400px", width = "100%")
```