---
always_allow_html: true
output:
  rmarkdown::github_document: default
  always_allow_html: default
  output:
  md_document:
    variant: gfm
knit: (function(inputFile, encoding) {
  rmarkdown::render(inputFile, encoding = encoding, output_dir = "../") })
date: "`r Sys.Date()`"
title: "SMITH-STUDY: Updated Data Processing; STRESS-Scale and Self-Efficacy Scale (EIPSES)"
---

```{r setup, include=FALSE}
base_dir <- "../" # where output should default to
base_url <- "" # keep as is
fig_path <- "_plot_images/" # where image output should go

knitr::opts_knit$set(base.dir = base_dir)
knitr::opts_chunk$set(fig.path = fig_path,
                      cache.path = 'cache/',  #log files go here
                      message=FALSE, warning=FALSE, cache = FALSE) 
```  

# 1. Setup & Data Ingestion
## 1.1 Libraries

```{r}
library(dplyr)       # Data manipulation
library(stringr)     # String manipulation
library(tidyr)       # Data pivoting
library(readr)       # Reading data
library(kableExtra)  # HTML Table formatting
library(ggplot2)     # Visualizations
library(ggpubr)      # Publication-ready plots
library(rstatix)     # T-tests and effect sizes
library(broom)       # Tidy statistical outputs
library(forcats)     # Factor manipulation
library(effectsize)  # effect size & omega squared

```

## 1.2 Load Data
### 1.2.0 !! Updated Data File:
- **UPDATED DATA FILE** - SEFF_Updated...
  - Both Surveys are connected to a single participant ID (Pre and Post) and collated in one spreadsheet
  - Will separate each into its own table, for clarity/checks and file write out. 
  - Rejoin and save data files for analysis scripts.

- **Structure**
  - Study_ID (Joined ID across pre and post, unique to participant)
  - Group Assignment (Control or Intervention)
  - Stress Scale: **PSS** - *Perceived Stress Scale* - 10 Questions
  - SEFF Scale: **EFF** - *Eearly Intervention Parent Self Efficacy* -  14 Questions
  - Columns as Pre --> Post 

### 1.2.1 Pre PSS
```{r}
# Read in raw dataset
ALL_DATA.raw <- read_csv("../_data/stress and SE ALL blind.csv", col_names = TRUE)

#Split to Pre and Post & Select only Study ID, Group Assignment and Questions:
pre.PSS.named <- 
  ALL_DATA.raw %>% 
    select("Study ID", "Group Assignment", 5:14) %>%
    rename("StudyID" = "Study ID", "Group_Assignment" = "Group Assignment") %>% 
    rename_with(~paste0("PSS.PRE_Q", (as.numeric(str_extract(., "\\d+$"))-4)), .cols = 3:12) %>% 
    mutate(across(contains("_Q"), tolower)) #ALL QUESTION TEXT tolower, as earlier data file had capitalization issues; "Agree / agree"

pre.PSS.named %>% kable(caption="STRESS - Pre Intervention Data") %>% 
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = F) %>%
  scroll_box(height = "400px", width = "100%")
```

### 1.2.2 Pre EFF
```{r}
#split to pre and select SEFF cols
pre.EFF.named <- 
  ALL_DATA.raw %>% 
    select("Study ID", "Group Assignment", 16:31) %>% 
    rename("StudyID" = "Study ID", "Group_Assignment" = "Group Assignment") %>% 
    rename_with(~paste0("EFF.PRE_Q", (as.numeric(str_extract(., "\\d+$"))-15)), .cols = 3:18) %>% 
    mutate(across(contains("_Q"), tolower)) #ALL QUESTION TEXT tolower, as earlier data file had capitalization issues; "Agree / agree"

pre.EFF.named %>% kable(caption="SEFF - Pre Intervention Data") %>% 
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = F) %>%    scroll_box(height = "400px", width = "100%")
```

### 1.2.3 Post PSS

```{r}
#Split Post & Select only Study ID, Group Assignment and Questions:
post.PSS.named <- 
  ALL_DATA.raw %>% 
  select("Study ID", "Group Assignment", 34:43) %>% 
    rename("StudyID" = "Study ID", "Group_Assignment" = "Group Assignment") %>% 
  rename_with(~paste0("PSS.POST_Q", (as.numeric(str_extract(., "\\d+$"))-33)), .cols = 3:12) %>% 
    mutate(across(contains("_Q"), tolower)) #ALL QUESTION TEXT tolower, as earlier data file had capitalization issues; "Agree / agree"

post.PSS.named %>% kable(caption="STRESS - Post Intervention Data") %>% 
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = F) %>%    scroll_box(height = "400px", width = "100%")
```

### 1.2.4 Post EFF
```{r}
#split to pre and select SEFF cols
post.EFF.named  <- 
  ALL_DATA.raw %>% 
    select("Study ID", "Group Assignment", 45:60) %>% 
    rename("StudyID" = "Study ID", "Group_Assignment" = "Group Assignment") %>% 
    rename_with(~paste0("EFF.POST_Q", (as.numeric(str_extract(., "\\d+$"))-44)), .cols = 3:18) %>% 
    mutate(across(contains("_Q"), tolower)) #ALL QUESTION TEXT tolower, as earlier data file had capitalization issues; "Agree / agree"

post.EFF.named %>% kable(caption="SEFF - Post Intervention Data") %>% 
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = F) %>%    scroll_box(height = "400px", width = "100%")
```

## 1.3 Joins & Item Checks

```{r}
#join tables
all.PSS.named <- left_join(pre.PSS.named, post.PSS.named, by = c("StudyID", "Group_Assignment"))

#check all unique strings in questions
PSS.uniques <- all.PSS.named %>%  
  select(contains("_Q")) %>% 
  summarise(across(everything(), ~list(unique(.)))) %>% 
    pivot_longer(everything(), names_to = "Question", values_to = "Unique_Responses") %>% 
    unnest(cols = c(Unique_Responses))

table(PSS.uniques$Unique_Responses)
```

```{r}
#join tables
all.EFF.named <- left_join(pre.EFF.named, post.EFF.named, by = c("StudyID", "Group_Assignment"))

#check all unique strings in questions
EFF.uniques <- 
  all.EFF.named %>% 
  select(contains("_Q")) %>% 
  summarise(across(everything(), ~list(unique(.)))) %>% 
    pivot_longer(everything(), names_to = "Question", values_to = "Unique_Responses") %>% 
    unnest(cols = c(Unique_Responses))

table(EFF.uniques$Unique_Responses)

```

### ------


# 2. Scoring & Data Transformation

```{r}
#all.PSS.named %>% glimpse()
```

## 2.1 PSS-STRESS:  Item Scoring and Reverse Coding

Scoring: PSS scores are obtained by reversing responses (e.g., 0 = 4, 1 = 3, 2 = 2, 3 = 1 & 4 = 0) to the four positively
stated items (items 4, 5, 7, & 8) and then summing across all scale items. A short 4 item scale can be made from
questions 2, 4, 5 and 10 of the PSS 10 item scale.

```{r}
PSS.mutated <- all.PSS.named %>% 
  # 1. Convert text responses to numeric scores
  mutate(across(contains("_Q"), ~case_when(
    . == "never" ~ 0,
    . == "almost never" ~ 1,
    . == "sometimes" ~ 2,
    . == "fairly often" ~ 3,
    . == "very often" ~ 4,
    TRUE ~ NA_real_
  ))) %>% 
  # 2. Reverse code specific items
  mutate(across(ends_with(c("Q4", "Q5", "Q7", "Q8")), ~case_when(
    . == 0 ~ 4,
    . == 1 ~ 3,
    . == 2 ~ 2,
    . == 3 ~ 1,
    . == 4 ~ 0,
    TRUE ~ NA_real_
  ))) %>% 
  rowwise() %>% #group rows (individual ID)
  # 3. Calculate total scores
  mutate(PRE_Stress.Total = sum(c_across(starts_with("PSS.PRE_Q")), na.rm = FALSE),
         POST_Stress.Total = sum(c_across(starts_with("PSS.POST_Q")), na.rm = FALSE),
         PRE_Stress.Mean = mean(c_across(starts_with("PSS.PRE_Q")), na.rm = FALSE),
         POST_Stress.Mean = mean(c_across(starts_with("PSS.POST_Q")), na.rm = FALSE),
         PRE_Stress.StDev = sd(c_across(starts_with("PSS.PRE_Q")), na.rm = FALSE),
         POST_Stress.StDev = sd(c_across(starts_with("PSS.POST_Q")), na.rm = FALSE)
         ) %>% 
    ungroup()

PSS.mutated %>% select(StudyID, Group_Assignment, contains("PRE_S"), contains("POST_S")) %>%
  kable(caption="PSS - Calculated Columns") %>% 
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = F) %>%    scroll_box(height = "400px", width = "100%")
```


## 2.2 Scoring Guidelines (Guimond et al., 2008):
    - Mean/Total scores are computed by aggregating all items of the specific scale.
    - Reverse Scored Items: 3, 5, 6, 8, 10, 12, 13, 15, 16.
    - Parent Outcome Expectations: 3, 4, 5, 6, 7, 8, 10, 12, 15, 16.
    - Parent Competence: 1, 2, 9, 14.

```{r}
#all.EFF.named %>% glimpse()
```

```{r}
EFF.mutated <- all.EFF.named %>% 
  # 1. Convert text to numeric Likert (1-7)
  mutate(across(contains("_Q"), ~case_match(.,
    "strongly disagree" ~ 1,
    "disagree" ~ 2,
    "somewhat disagree" ~ 3,
    "neutral" ~ 4,
    "somewhat agree" ~ 5,
    "agree" ~ 6,
    "strongly agree" ~ 7,
    .default = NA_real_
  ))) %>% 
  
  # 2. Reverse Code specific items
  mutate(across(ends_with(c("Q3", "Q5", "Q6", "Q8", "Q10", "Q12", "Q13", "Q15", "Q16")), 
                ~ 8 - .)) %>% #  For 1-7 reverse coding (8 - 1 = 7, 8 - 7 = 1)

  # 3. Mean replace NAs
  mutate(across(contains("_Q"), ~ifelse(is.na(.), mean(., na.rm = TRUE), .))) %>%
  
  # 4. Calculate Composite Totals
  mutate(
    # PRE Composites
    PRE_Parent_Outcome_Expectations.Sum = rowSums(select(., ends_with(c("PRE_Q3", "PRE_Q4", "PRE_Q5", "PRE_Q6", "PRE_Q7", "PRE_Q8", "PRE_Q10", "PRE_Q12", "PRE_Q15", "PRE_Q16"))), na.rm = TRUE),
    PRE_Parent_Competence.Sum = rowSums(select(., ends_with(c("PRE_Q1", "PRE_Q2", "PRE_Q9", "PRE_Q14"))), na.rm = TRUE),
    PRE_Overall_SEFF.Sum = rowSums(select(., contains("PRE_Q")), na.rm = TRUE),
    
    # POST Composites
    POST_Parent_Outcome_Expectations.Sum = rowSums(select(., ends_with(c("POST_Q3", "POST_Q4", "POST_Q5", "POST_Q6", "POST_Q7", "POST_Q8", "POST_Q10", "POST_Q12", "POST_Q15", "POST_Q16"))), na.rm = TRUE),
    POST_Parent_Competence.Sum = rowSums(select(., ends_with(c("POST_Q1", "POST_Q2", "POST_Q9", "POST_Q14"))), na.rm = TRUE),
    POST_Overall_SEFF.Sum = rowSums(select(., contains("POST_Q")), na.rm = TRUE)
  ) %>% 
  # 5. Calculate Composite MEANS
  mutate(
    # PRE Composites
    PRE_Parent_Outcome_Expectations.Mean = rowMeans(select(., ends_with(c("PRE_Q3", "PRE_Q4", "PRE_Q5", "PRE_Q6", "PRE_Q7", "PRE_Q8", "PRE_Q10", "PRE_Q12", "PRE_Q15", "PRE_Q16"))), na.rm = TRUE),
    PRE_Parent_Competence.Mean = rowMeans(select(., ends_with(c("PRE_Q1", "PRE_Q2", "PRE_Q9", "PRE_Q14"))), na.rm = TRUE),
    PRE_Overall_SEFF.Mean = rowMeans(select(., contains("PRE_Q")), na.rm = TRUE),
    
    # POST Composites
    POST_Parent_Outcome_Expectations.Mean = rowMeans(select(., ends_with(c("POST_Q3", "POST_Q4", "POST_Q5", "POST_Q6", "POST_Q7", "POST_Q8", "POST_Q10", "POST_Q12", "POST_Q15", "POST_Q16"))), na.rm = TRUE),
    POST_Parent_Competence.Mean = rowMeans(select(., ends_with(c("POST_Q1", "POST_Q2", "POST_Q9", "POST_Q14"))), na.rm = TRUE),
    POST_Overall_SEFF.Mean = rowMeans(select(., contains("POST_Q")), na.rm = TRUE)
  ) #%>% 
  #mutate(Filter_Diff = POST_Overall_SEFF.Mean - PRE_Overall_SEFF.Mean)

if(is.na(EFF.mutated) %>% sum() > 0){
  # Check for any remaining NAs in the final scored dataset
  print("ERROR - NA OR MISSING DATA")
}
```


```{r}
EFF.mutated %>% select(StudyID, Group_Assignment, contains(c("Total","Mean"))) %>%
  kable(caption="EFF - Calculated Columns") %>% 
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = F) %>%    scroll_box(height = "400px", width = "100%")
```


# 3. Save Analaysis Tables and Write File for Download/Use

```{r}

EFF.scored <- EFF.mutated
PSS.scored <- PSS.mutated

analytic.files <- list(
  "EFF.scored" = EFF.scored,
  "PSS.Scored" = PSS.scored
  )

readr::write_rds(
  analytic.files,
  path = "../_output/analytic_files_scored_surveys.Rds" )

```
