---
always_allow_html: true
output:
  rmarkdown::github_document: default
  always_allow_html: default
  output:
  md_document:
    variant: gfm
knit: (function(inputFile, encoding) {
  rmarkdown::render(inputFile, encoding = encoding, output_dir = "../") })
date: "`r Sys.Date()`"
title: "Updated Analysis: STRESS (Perceived Stress Scale)"
---

```{r setup, include=FALSE}
base_dir <- "../" # where output should default to
base_url <- "" # keep as is
fig_path <- "_plot_images/" # where image output should go

knitr::opts_knit$set(base.dir = base_dir)
knitr::opts_chunk$set(fig.path = fig_path,
                      cache.path = 'cache/',  #log files go here
                      message=FALSE, warning=FALSE, cache = FALSE) 
```  


```{r include=FALSE}
# .base_dir <- "../" 
# .fig_path <- "_plot_images/" 
# 
# knitr::opts_knit$set(base.dir = .base_dir)
# knitr::opts_chunk$set(
#   fig.path = .fig_path,
#   cache.path = 'cache/', 
#   cache = FALSE,
#   message = FALSE, 
#   warning = FALSE,
#   echo = TRUE
# )

```

# 0. Setup & Data Ingestion
## 0.1 Libraries

```{r}
library(dplyr)       # Data manipulation
library(stringr)     # String manipulation
library(tidyr)       # Data pivoting
library(readr)       # Reading data
library(kableExtra)  # HTML Table formatting
library(ggplot2)     # Visualizations
library(ggpubr)      # Publication-ready plots
library(rstatix)     # T-tests and effect sizes
library(broom)       # Tidy statistical outputs
library(forcats)     # Factor manipulation
library(effectsize)  # effect size & omega squared

#install.packages("skimr")
#install.packages("gtExtras") # testing: used for quick summaries and viz
```

## 0.2 Load Data
### 0.2.1 !! Updated Data File:
- **UPDATED DATA FILE** - SEFF_Updated...
  - Both Surveys are connected to a single participant ID (Pre and Post) and collated in one spreadsheet
  - Transformed to numeric, scored, and saved in [Autism_Doula_Study_data_ingestion.Rmd](SEFF_Updated_Data_File_08-2026.Rmd)

- **Structure**
  - Study_ID (Joined ID across pre and post, unique to participant)
  - Group Assignment (Control or Intervention)
  - Stress Scale: **PSS** - *Perceived Stress Scale* - 10 Questions
  - SEFF Scale: **EFF** - *Early Intervention Parent Self Efficacy* -  14 Questions
  - Columns as Pre --> Post 


## 0.3 Load Analaysis Tables
```{r}
analytic.files <- read_rds("../_output/analytic_files_scored_surveys.Rds")

EFF.scored <- analytic.files$EFF.scored
PSS.scored <- analytic.files$PSS.Scored
```

```{r}
# library(gtExtras)
# 
# PSS.scored %>% 
#   mutate(StudyID = factor(StudyID), Group_Assignment = factor(Group_Assignment)) %>%  
#   select(StudyID, Group_Assignment, contains("PRE_S"), contains("POST_S")) %>% 
#   gt_plt_summary()

```

# 1. PSS
```{r}
PSS.scored %>% select(StudyID, Group_Assignment, contains("PRE_S"), contains("POST_S")) %>%
  kable(caption="PSS - Calculated Columns") %>% 
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = F) %>%    scroll_box(height = "400px", width = "100%")
```
## 1.1 Descriptiive Table
```{r}
PSS.scored %>%
  summarise(.by = Group_Assignment, 
            N = n(),
            Mean_PRE = mean(PRE_Stress.Total, na.rm = TRUE),
            SD_PRE = sd(PRE_Stress.Total, na.rm = TRUE),
            Mean_POST = mean(POST_Stress.Total, na.rm = TRUE),
            SD_POST = sd(POST_Stress.Total, na.rm = TRUE)
            ) %>%
  kable(digits = 2,
        caption = "Summary Table - Group Means and St.Deviations") %>%
  kable_styling(full_width = F)
```

## 1.2 Norm Table from PSS Original Study Paper
  - Taken from the Paper, 
  - Using **Total** Scores (Means for Demographic Groups)
  - Compare 
    - Study: Intervention (24) M = 16.4, SD = 6.86. (Pre Intervention)
    - Study: Control (7) M = 14.7, SD = 7.91
    - Race: Black (176) M = 14.7, SD = 7.2

| Category | N   | Mean | S.D. |
|---------|-----|------|------|
| Gender<br>Male     | 926 | 12.1 | 5.9  |
| Gender<br>Female   | 1406| 13.7 | 6.6  |
| Age<br>18-29       | 645 | 14.2 | 6.2  |
| Age<br>30-44       | 750 | 13.0 | 6.2  |
| Age<br>45-54       | 285 | 12.6 | 6.1  |
| Age<br>55-64       | 282 | 11.9 | 6.9  |
| Age<br>65 & older  | 296 | 12.0 | 6.3  |
| Race<br>White      | 1924| 12.8 | 6.2  |
| Race<br>Hispanic   | 98  | 14.0 | 6.9  |
| Race<br>Black      | 176 | 14.7 | 7.2  |
| Race<br>Other Minority |50 | 14.1 | 5.0 |


## 1.2 PSS - Scale Validation


### 1.2 Item-Level Distribution (Dot Plot)
```{r}
PSS.scored %>% 
  filter(Group_Assignment == "Intervention") %>% 
  pivot_longer(cols = contains("_Q"), names_to = "Item", values_to = "Value") %>% 
  mutate(
    Condition = if_else(str_detect(Item, "PRE_"), "Pre-Test", "Post-Test"),
    Condition = factor(Condition, levels = c("Pre-Test", "Post-Test")),
    Q_Num = as.numeric(str_extract(Item, "\\d+")),
    Q_Order = factor(Q_Num, levels = 1:10)
    #Composite = if_else(Q_Num %in% c(1, 2, 9, 14), "Parent Competence", "Outcome Expectations")
  ) %>% 
  filter(!is.na(Value)) %>% 
  ggplot(aes(x = Value, y = fct_rev(Q_Order)))+#, color = Composite)) +
  geom_jitter(position = position_jitter(width = 0.2, height = 0.1), alpha = 0.5) +
  stat_summary(fun = "mean", geom = "point", color = "orange", size = 2, shape = 18) +
  facet_wrap(~Condition) +
  scale_x_continuous(breaks = 0:4) +
  theme_minimal() +
  theme(legend.position = "top") +
  labs(
    title = "Item-Level Distributions (Likert Scale 0 - 4 | Never - Very Often)",
    subtitle = "Black diamond (♦) represents the item mean.",
    y = "Question Number",
    x = "Likert Score"
  )
```


## 2. Test Internal Consistency (Cronbach's Alpha)
- Here we will calculate Cronbach's alpha for each scale to check for internal consistency. 
- A general rule of thumb is that an alpha of > 0.7 is acceptable.

```{r}

pre_alpha <- 
  PSS.scored %>% 
  filter(Group_Assignment == "Intervention") %>% 
  select(contains("PRE_Q")) %>% 
  psych::alpha()

post_alpha <- 
  PSS.scored %>% 
  filter(Group_Assignment == "Intervention") %>% 
  select(contains("POST_Q")) %>% 
  psych::alpha()


## Dataframe of the Cronbach Alpha Tests and Values
alpha_values <- data.frame(
  Scale = c("PRE", "POST"),
  Cronbach_Alpha = c(pre_alpha$total$raw_alpha,
                     post_alpha$total$raw_alpha)
)

alpha_values %>% 
  kableExtra::kable(digits=3,
                    caption = "Intervention Group: Cronbach Alpha Values") %>%
  kableExtra::kable_styling(bootstrap_options = c("striped", "hover", "condensed"), 
                            full_width = F)
```

#3. Descriptive Statistics

## 3.1 Descriptive Table
```{r}
PSS.scored %>%
  summarise(.by = Group_Assignment, 
            N = n(),
            Mean_PRE = mean(PRE_Stress.Total, na.rm = TRUE),
            SD_PRE = sd(PRE_Stress.Total, na.rm = TRUE),
            Mean_POST = mean(POST_Stress.Total, na.rm = TRUE),
            SD_POST = sd(POST_Stress.Total, na.rm = TRUE)
            ) %>%
  kable(digits = 2,
        caption = "Summary Table - Group Means and St.Deviations") %>%
  kable_styling(full_width = F)
```

### 3.2.1 Outlier Identification
```{r}
PSS.check <- PSS.scored %>% mutate(Filter_Diff = POST_Stress.Mean - PRE_Stress.Mean)
#Control doesn't have POST scores, so they are ignored via NA-REMOVE  
f.mn <- mean(PSS.check$Filter_Diff, na.rm = TRUE)
f.sd <- sd(PSS.check$Filter_Diff, na.rm = TRUE)

PSS.check <- PSS.check %>% 
  mutate(
    Filter_Criterion = ifelse(Filter_Diff > (f.mn + 2*f.sd) | Filter_Diff < (f.mn - 2*f.sd), "Exclude", "Include")
  )

PSS.check %>% 
  filter(Group_Assignment == "Intervention") %>% 
  ggplot(aes(x=Filter_Diff, fill = Filter_Criterion)) + geom_histogram(bins=10, color="black") + 
  labs(title="Distribution of Pre/Post Differences in Overall SEFF Scores", x="Difference (Post - Pre)", y="Count") +
  theme_minimal()
  
```

# 3.3 Run Paired T-Tests

## Are the groups different initially?
```{r}
PSS.check %>% rstatix::t_test(PRE_Stress.Total ~ Group_Assignment, paired = FALSE, alternative = "two.sided", conf.level = 0.1)

```

## Does the intervention group change pre to post test?

```{r}
PSS.long <- PSS.check %>% 
  #filter(Filter_Criterion == "Include") %>%
  select(StudyID, Group_Assignment, contains("Mean")) %>%  
  pivot_longer(cols = contains("Stress"), names_to = "Composite", values_to = "Score") %>% 
  mutate(
    Timepoint = if_else(str_detect(Composite, "PRE_"), "Pre-Test", "Post-Test"),
    Timepoint = factor(Timepoint, levels = c("Pre-Test", "Post-Test")),
  )

# Run Paired T-Tests
PSS.t_test_results <- PSS.long %>%
  rstatix::pairwise_t_test(Score ~ Timepoint, paired = TRUE, detailed = TRUE, conf.level = 0.1, alternative = "less", pool.sd=TRUE)
  #adjust_pvalue(method = "holm") %>%
  #add_significance("p.adj")

PSS.t_test_results


```

## One Sample Test (Against Control & Norm Mean)
- Control Group (7) is too small to use as a reference group for the T-Test (independent sample t-test, Intervention POST vs Control PRE)
- But Control mean (14.7) is identical to the Normed Mean for the Race: Black Demographic
- Can test with one sample t-test against the norm value for the group.
$$ Post Stress Total_I == 14.7 $$
```{r}
PSS.check %>% 
  #filter(Filter_Criterion == "Include") %>%
  select(StudyID, Group_Assignment, POST_Stress.Total) %>%  
  filter(Group_Assignment == "Intervention") %>% 
  rstatix::t_test(POST_Stress.Total ~ 1, mu = 14.7) 
```
- Post Stress Sample is NOT significantly different from the normed (race group) or control. 
- Indication that the intervention may have reduced scores to where they are at least close to normal, although not large enough to account for **starting differences** (paired test) for the individuals levels. 


# 4. Plots

## Pre, Post and Control Boxplots
```{r}
PSS.pre_boxplot <- 
  PSS.scored %>% 
  #select(StudyID, Group_Assignment, contains("Total"), contains("Mean")) %>% 
  ggplot(aes(x = Group_Assignment, y = PRE_Stress.Mean))+
  geom_boxplot()+
  geom_jitter()

PSS.post_boxplot <- 
  PSS.scored %>% 
  #select(StudyID, Group_Assignment, contains("Total"), contains("Mean")) %>% 
  ggplot(aes(x = Group_Assignment, y = POST_Stress.Mean))+
  geom_boxplot()+
  geom_jitter()

library(patchwork)

PSS.pre_boxplot + PSS.post_boxplot + plot_annotation(title = "PSS - Pre/Post Stress Scores by Group Assignment")

```
```{r}
PSS.check %>% 
  select(StudyID, Group_Assignment, contains("Mean"), Filter_Criterion) %>%  
  pivot_longer(cols = contains("Stress"), names_to = "Item", values_to = "Value") %>% 
  mutate(
    Condition = if_else(str_detect(Item, "PRE_"), "Pre-Test", "Post-Test"),
    Condition = factor(Condition, levels = c("Pre-Test", "Post-Test")),
  ) %>% 
  ggplot(aes(x = fct_rev(Item), y = Value, color = Group_Assignment, fill = Group_Assignment)) +
  geom_boxplot(alpha = .5) +
  geom_jitter(aes(shape = Filter_Criterion), position = position_jitter(width = 0.2, height = 0.1), alpha = 0.8)

```


## 4.1 Change in Total Scores (Arrow Plot)
```{r}
PSS.scored %>% 
  arrange(PRE_Stress.Mean) %>% 
  mutate(
    Rank_Order = seq_along(PRE_Stress.Mean),
    Pct_Order = percent_rank(PRE_Stress.Mean)
  ) %>% 
  ggplot(aes(x = Rank_Order)) +
  # 1. Add the arrows (vertical segments)
  geom_segment(aes(
    xend = Rank_Order, 
    y = PRE_Stress.Mean, 
    yend = POST_Stress.Mean
  ), 
  arrow = arrow(length = unit(.35, "cm")), # Adds the arrow head
  color = "gray", 
  alpha = 0.5) +
  
  # 2. Add the PRE points
  geom_point(aes(y = PRE_Stress.Mean, color = "PRE_Stress.Mean")) +
  
  # 3. Add the POST points
  geom_point(aes(y = POST_Stress.Mean, color = "POST_Stress.Mean")) +
  
  # Formatting
  scale_color_manual(values = c("PRE_Stress.Mean" = "red", "POST_Stress.Mean" = "blue")) +
  labs(
    title = "Individual Stress Score Change (Ranked by PRE Score)",
    x = "Participant Rank (Low to High PRE Stress)",
    y = "Total Stress Score",
    color = "Timepoint"
  ) +
  theme_minimal()
```

