---
always_allow_html: true
output:
  rmarkdown::github_document: default
  always_allow_html: default
  output:
  md_document:
    variant: gfm
knit: (function(inputFile, encoding) {
  rmarkdown::render(inputFile, encoding = encoding, output_dir = "../") })
date: "`r Sys.Date()`"
title: "Updated_Analysis; Self-Efficacy (EIPSES)"
---

```{r setup, include=FALSE}
base_dir <- "../" # where output should default to
base_url <- "" # keep as is
fig_path <- "_plot_images/" # where image output should go

knitr::opts_knit$set(base.dir = base_dir)
knitr::opts_chunk$set(fig.path = fig_path,
                      cache.path = 'cache/',  #log files go here
                      message=FALSE, warning=FALSE, cache = FALSE) 
```  


```{r include=FALSE}
# .base_dir <- "../" 
# .fig_path <- "_plot_images/" 
# 
# knitr::opts_knit$set(base.dir = .base_dir)
# knitr::opts_chunk$set(
#   fig.path = .fig_path,
#   cache.path = 'cache/', 
#   cache = FALSE,
#   message = FALSE, 
#   warning = FALSE,
#   echo = TRUE
# )

```

# 0. Setup & Data Ingestion
## 0.1 Libraries

```{r}
library(dplyr)       # Data manipulation
library(stringr)     # String manipulation
library(tidyr)       # Data pivoting
library(readr)       # Reading data
library(kableExtra)  # HTML Table formatting
library(ggplot2)     # Visualizations
library(ggpubr)      # Publication-ready plots
library(rstatix)     # T-tests and effect sizes
library(broom)       # Tidy statistical outputs
library(forcats)     # Factor manipulation
library(effectsize)  # effect size & omega squared

#install.packages("skimr")
#install.packages("gtExtras") # testing: used for quick summaries and viz
```

## 0.2 Load Data
### 0.2.1 !! Updated Data File:
- **UPDATED DATA FILE** - SEFF_Updated...
  - Both Surveys are connected to a single participant ID (Pre and Post) and collated in one spreadsheet
  - Transformed to numeric, scored, and saved in [Autism_Doula_Study_data_ingestion.Rmd](SEFF_Updated_Data_File_08-2026.Rmd)

- **Structure**
  - Study_ID (Joined ID across pre and post, unique to participant)
  - Group Assignment (Control or Intervention)
  - Self-Efficacy Scale: **PSS** - *Perceived Self-Efficacy Scale* - 10 Questions
  - SEFF Scale: **EFF** - *Early Intervention Parent Self Efficacy* -  14 Questions
  - Columns as Pre --> Post 


## 0.3 Load Analaysis Tables
```{r}
analytic.files <- read_rds("../_output/analytic_files_scored_surveys.Rds")

EFF.scored <- analytic.files$EFF.scored
PSS.scored <- analytic.files$PSS.Scored
```

```{r}
# library(gtExtras)
# 
# 
# EFF.scored %>% 
#   mutate(StudyID = factor(StudyID), Group_Assignment = factor(Group_Assignment)) %>%  
#   select(StudyID, Group_Assignment, contains(c("Sum","Mean"))) %>% 
#   gt_plt_summary()

```


 --- OUTLINE ---
Descriptive and Participant Checks
(Missing Data / NAs)
Basic Histo/Frequency Etc...
Difference Scores
Dot Plots

--- Stats ---
Scale Validation

--- Tests --- 
Pre & Post

--- Models ---
Pre/Post Linear Mixed Models Etc...

# 1. EFF
```{r}
EFF.scored %>% select(StudyID, Group_Assignment, contains(c("Sum","Mean"))) %>%
  kable(caption="EFF - Calculated Columns") %>% 
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = F) %>%    scroll_box(height = "400px", width = "100%")
```
## 1.1 Descriptive Table
```{r}
library(tableone)
EFF.t1 <- 
tableone::CreateTableOne(
  data = EFF.scored,
  vars = c("PRE_Overall_SEFF.Mean", "POST_Overall_SEFF.Mean","PRE_Parent_Outcome_Expectations.Mean", "POST_Parent_Outcome_Expectations.Mean","PRE_Parent_Competence.Mean", "POST_Parent_Competence.Mean"),
  strata = "Group_Assignment",
  factorVars = c("Group_Assignment"),
  testNormal = "oneway.test"
) 
EFF.t1 %>%   print(showAllLevels = TRUE, quote = FALSE, noSpaces = TRUE) %>% 
  kable(caption="EFF - Descriptive Table") %>% 
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = F) %>%    scroll_box(height = "400px", width = "100%")
```

```{r}
EFF.long <- EFF.scored %>%
  select(StudyID, Group_Assignment, ends_with(c(".Sum",".Mean"))) %>%
  pivot_longer(cols = ends_with(c(".Sum",".Mean")), names_to = "Variable", values_to = "Score") %>%
  separate(Variable, into = c("Condition", "Composite_Type"), sep = "_", extra = "merge") %>%
  separate(Composite_Type, into = c("Composite_Type", "Calculation"), sep = "\\.", extra = "merge")


EFF.int.desc <- 
  EFF.long %>% 
    filter(Group_Assignment == "Intervention") %>%
    filter(Calculation == "Mean") %>%
    group_by(Group_Assignment, Condition, Composite_Type) %>% 
    summarize(
    N = n(),
    Missing = sum(Score==0),
    Mean = mean(Score, na.rm=TRUE),
    SD = sd(Score, na.rm=TRUE),
    Min = min(Score, na.rm=TRUE),
    Max = max(Score, na.rm=TRUE),
    .groups = "drop"
  )

  
EFF.int.desc %>%
 kable(digits = 2,
       caption = "Summary Table - Group Means and St.Deviations") %>%
 kable_styling(full_width = F)
```

## 1.2 Norm Table from EFF Original Study Paper
  - Excerpted Rows from Guimond et al. 2008 
  - Using **Mean** Scores (Means for Demographic Groups)
  - Grouped by Child demographic, Parents are still the survey subject. 
  
| Group             | N                         | Total EIPSES (M, SD) | Parent Outcome Expectations (M, SD) | Parent Competence (M, SD) |
|-------------------|---------------------------|----------------------|------------------------|--------------------|
| Child gender<br>Boys                | 72      | 5.75 (.64)           | 5.68 (.79)             | 5.90 (.79)         |
| Child gender<br>Girls               | 45      | 5.94 (.65)           | 5.91 (.73)             | 6.02 (.73)         |
| Ethnicity<br>White                  | 77      | 5.92 (.59)           | 5.93 (.64)             | 5.89 (.74)         |
| Ethnicity<br>Hispanic               | 27      | 5.65 (.79)           | 5.44 (1.03)            | 6.07 (.74)         |
| Ethnicity<br>Other                  | 13      | 5.65 (.55)           | 5.51 (.64)             | 6.02 (.69)         |
| Child disorder severity<br>Mild     | 46      | 5.97 (.63)           | 5.92 (.75)             | 6.11 (.70)         |
| Child disorder severity<br>Modereate | 44     | 5.76 (.64)           | 5.71 (.75)             | 5.89 (.79)         |
| Child disorder severity<br>Severe   | 11      | 6.05 (.54)           | 6.05 (.56)             | 6.05 (.76)         |


## 1.2 EFF - Scale Validation

## 2. Test Internal Consistency (Cronbach's Alpha)
- Here we will calculate Cronbach's alpha for each scale to check for internal consistency. 
- A general rule of thumb is that an alpha of > 0.7 is acceptable.
- **PRE Group** is just under .7 ~ should check for consistency/outliers. 

```{r}

pre_alpha <- 
  EFF.scored %>% 
  filter(Group_Assignment == "Intervention") %>% 
  select(contains("EFF.PRE_Q")) %>% 
  psych::alpha()

post_alpha <- 
  EFF.scored %>% 
  filter(Group_Assignment == "Intervention") %>% 
  select(contains("EFF.POST_Q")) %>% 
  psych::alpha()


## Dataframe of the Cronbach Alpha Tests and Values
alpha_values <- data.frame(
  Scale = c("PRE", "POST"),
  Cronbach_Alpha = c(pre_alpha$total$raw_alpha,
                     post_alpha$total$raw_alpha)
)

alpha_values %>% 
  kableExtra::kable(digits=3,
                    caption = "Intervention Group: Cronbach Alpha Values") %>%
  kableExtra::kable_styling(bootstrap_options = c("striped", "hover", "condensed"), 
                            full_width = F)
```


### 1.2 Item-Level Distribution (Dot Plot)
```{r}
EFF.scored %>% 
  filter(Group_Assignment == "Intervention") %>% 
pivot_longer(cols = contains("_Q"), names_to = "Item", values_to = "Value") %>% 
  mutate(
    Condition = if_else(str_detect(Item, "PRE_"), "Pre-Test", "Post-Test"),
    Condition = factor(Condition, levels = c("Pre-Test", "Post-Test")),
    Q_Num = as.numeric(str_extract(Item, "\\d+")),
    Q_Order = factor(Q_Num, levels = 1:16),
    Composite = if_else(Q_Num %in% c(1, 2, 9, 14), "Parent Competence", "Outcome Expectations")
  ) %>% 
  filter(!is.na(Value)) %>% 
  ggplot(aes(x = Value, y = fct_rev(Q_Order), color = Composite)) +
  geom_jitter(position = position_jitter(width = 0.2, height = 0.1), alpha = 0.5) +
  stat_summary(fun = "mean", geom = "point", color = "black", size = 2, shape = 18) +
  facet_wrap(~Condition) +
  scale_x_continuous(breaks = 1:7) +
  theme_minimal() +
  theme(legend.position = "top") +
  labs(
    title = "Intervention Group - Item-Level Distributions (Likert 1-7)",
    subtitle = "Black diamond (♦) represents the item mean.",
    y = "Question Number",
    x = "Likert Score"
  )
```

#2. Descriptive Statistics

## 2.1 Descriptive Table
```{r}
EFF.scored %>%
  summarise(.by = Group_Assignment, 
            N = n(),
            Mean_PRE = mean(PRE_Overall_SEFF.Sum, na.rm = TRUE),
            SD_PRE = sd(PRE_Overall_SEFF.Sum, na.rm = TRUE),
            Mean_POST = mean(POST_Overall_SEFF.Sum, na.rm = TRUE),
            SD_POST = sd(POST_Overall_SEFF.Sum, na.rm = TRUE)
            ) %>%
  kable(digits = 2,
        caption = "Summary Table - Group Totals and St.Deviations") %>%
  kable_styling(full_width = F)
```

### 2.2 Outlier Identification
```{r}
EFF.check <- EFF.scored %>% mutate(Filter_Diff = POST_Overall_SEFF.Mean - PRE_Overall_SEFF.Mean,
                                   Filter_Diff = ifelse(Group_Assignment == "Control", NA, Filter_Diff))
#Control scores changed to NA and then ignored
f.mn <- mean(EFF.check$Filter_Diff, na.rm = TRUE)
f.sd <- sd(EFF.check$Filter_Diff, na.rm = TRUE)

EFF.check <- EFF.check %>% 
  mutate(
    Filter_Criterion = ifelse(Filter_Diff > (f.mn + 2*f.sd) | Filter_Diff < (f.mn - 2*f.sd), "Exclude", "Include")
  )

EFF.check %>% 
  filter(Group_Assignment == "Intervention") %>% 
  ggplot(aes(x=Filter_Diff, fill = Filter_Criterion)) + geom_histogram(bins=10, color="black") + 
  labs(title="Distribution of Pre/Post Differences in Overall SEFF Scores", x="Difference (Post - Pre)", y="Count") +
  theme_minimal()
  
```

# 3. Statistical Tests 

## 3.1 Equal Variances
 - Standard assumption checks; given the small sample size, these aren't too meaningful, especially as we are going to be running **pairwise** tests (same participants in the intervention group). 
 - Check the assumptions first: normality (shapiro_test() 
 - Paired Tests (Student test) equal variances (levene_test() / var.test()). 
  - If normality fails, switch to the Wilcoxon test, adjust variance, check outliers... 
```{r}
EFF.check %>% filter(Group_Assignment == "Intervention") %>% rstatix::shapiro_test(PRE_Overall_SEFF.Mean, POST_Overall_SEFF.Mean) %>% 
  mutate("Normality" = ifelse(p < 0.05, "Reject Null (Non-Normal)", "Fail to Reject Null (Normal)"))
```


```{r}
EFF.check %>% 
  filter(Group_Assignment == "Control") %>% 
  rstatix::shapiro_test(PRE_Overall_SEFF.Mean) %>% 
  mutate("Normality" = ifelse(p < 0.05, "Reject Null (Non-Normal)", "Fail to Reject Null (Normal)"))
```


```{r}
EFF.check %>% rstatix::levene_test(PRE_Overall_SEFF.Mean ~ Group_Assignment) %>% 
  mutate("Equal Variance" = ifelse(p < 0.05, "Reject Null (Adjust Test)", "Fail to Reject Null (Standard Test)"))
```
**fail to reject null; standard t-test is fine**


```{r}
EFF.check %>% rstatix::levene_test(POST_Overall_SEFF.Mean ~ Group_Assignment) %>% 
  mutate("Equal Variance" = ifelse(p < 0.05, "Reject Null (Adjust Test)", "Fail to Reject Null (Standard Test)"))
## reject null: use POOLED variance (or Welch's) ; likely to be the case given that Inervention POST was non-normal. 

```


# 3.3 Run Paired T-Tests

## Are the groups different initially?
```{r}
EFF.check %>% 
  rstatix::t_test(PRE_Overall_SEFF.Mean ~ Group_Assignment, 
                  paired = FALSE, # Intervention v Control
                  alternative = "two.sided", #any difference 
                  conf.level = 0.1, 
                  var.equal = TRUE #levene reject null
                  ) %>% 
  mutate("Inference" = ifelse(p < 0.05, "Reject Null - Groups are diff", "Fail to Reject Null (groups same)"))
```
- Not reading too much in yet as control sample is so small; (n=7). 
- With no difference here, we will see if Pre --> Post Intervention makes a difference *greater Efficacy*
- Compare Changes in intervention group. 
- Compare Post Intervention to Control (if size allows...) 
- `rstatix` used; mirrors the calculations for SPSS


## T-Test - Intervention Group, Pre v Post

```{r}
EFF.long <- EFF.check %>%
  select(StudyID, Group_Assignment, Filter_Criterion, ends_with(c(".Sum",".Mean"))) %>%
  pivot_longer(cols = ends_with(c(".Sum",".Mean")), names_to = "Variable", values_to = "Score") %>%
  separate(Variable, into = c("Condition", "Composite_Type"), sep = "_", extra = "merge") %>%
  separate(Composite_Type, into = c("Composite_Type", "Calculation"), sep = "\\.", extra = "merge") %>% 
  mutate(Condition.f = factor(Condition, levels = c("PRE", "POST")))

library(rstatix)

# Run Paired T-Tests
t_test_results <- EFF.long %>%
  filter(Group_Assignment == "Intervention") %>% 
  filter(Calculation == "Mean") %>%
  group_by(Composite_Type) %>%
  rstatix::pairwise_t_test(Score ~ Condition.f, paired = TRUE, detailed = TRUE, conf.level = 0.1, alternative = "greater", pool.sd=TRUE)
  #adjust_pvalue(method = "holm") %>%
  #add_significance("p.adj")

t_test_results

```

## Wilcox Test

```{r}
EFF.long %>% 
  filter(Group_Assignment == "Intervention") %>% 
  filter(Calculation == "Mean") %>%
  group_by(Composite_Type) %>% 
  rstatix::wilcox_test(Score ~ Condition.f, paired = TRUE, alternative = "greater", conf.level = 0.1)
```

## Interpretation
- Groups scores (Control & Intervention) are the same **PRE-INTERVENTION**
- Intervention Group scores do NOT significantly increase (from Pre to Post intervention)

```{r}
# # Calculate Cohen's d for Effect Size
# cohens_d_results <- EFF.long %>%
#   filter(Group_Assignment == "Intervention") %>% 
#   filter(Calculation == "Mean") %>%
#   group_by(Composite_Type) %>%
#   rstatix::cohens_d(Score ~ Condition, paired = TRUE)
# 
# # Combine for final table
# t_test_final <- t_test_results %>%
#   left_join(select(cohens_d_results, Composite_Type, effsize, magnitude), by = "Composite_Type") %>%
#   select(Composite_Type, statistic, df, p, p.adj, p.adj.signif, effsize, magnitude)
# 
# t_test_final %>% 
#   kable(caption = "Paired T-Test Results (Mean: Pre vs. Post)", digits = 3) %>% 
#   kable_styling(bootstrap_options = c("striped", "hover"), full_width = T)
```




```{r}
## Does the intervention group change pre to post test?
# EFF.test.long <- EFF.check %>% 
#   filter(Filter_Criterion == "Include") %>%
#   filter(Group_Assignment == "Intervention") %>% 
#   select(StudyID, contains("Overall_SEFF.Mean")) %>%  
#   pivot_longer(cols = contains(c("Overall", "Parent")), names_to = "Composite", values_to = "Score") %>% 
#   mutate(
#     Timepoint = if_else(str_detect(Composite, "PRE_"), "Pre-Test", "Post-Test"),
#     Timepoint = factor(Timepoint, levels = c("Pre-Test", "Post-Test")),
#   )
# 
# EFF.test.long
# 
# # Run Paired T-Tests
# EFF.t_test_results <- EFF.test.long %>%
#   rstatix::pairwise_t_test(Score ~ Timepoint, paired = TRUE, detailed = TRUE, conf.level = 0.1, alternative = "greater", pool.sd=TRUE)
#   #adjust_pvalue(method = "holm") %>%
#   #add_significance("p.adj")
# 
# EFF.t_test_results

```


```{r}
# EFF.check %>% 
#   filter(Filter_Criterion == "Include") %>%
#   filter(Group_Assignment == "Intervention") %>% 
#   select(StudyID, contains("Parent_Competence.Mean")) %>%  
#   pivot_longer(cols = contains(c("Overall", "Parent")), names_to = "Composite", values_to = "Score") %>% 
#   mutate(
#     Timepoint = if_else(str_detect(Composite, "PRE_"), "Pre-Test", "Post-Test"),
#     Timepoint = factor(Timepoint, levels = c("Pre-Test", "Post-Test")),
#   ) %>% 
# 
# # Run Paired T-Tests
# #EFF.t_test_results <- EFF.test.long %>%
#   rstatix::pairwise_t_test(Score ~ Timepoint, paired = TRUE, detailed = TRUE, conf.level = 0.1, alternative = "greater", pool.sd=TRUE)
#   #adjust_pvalue(method = "holm") %>%
#   #add_significance("p.adj")
# 
# EFF.t_test_results
```


## One Sample Test (Against Control & Normed Mean)
- Normed Mean from EIPSES table, 5.47 (Race: Other)

```{r}
EFF.check %>% 
  #filter(Filter_Criterion == "Include") %>%
  select(StudyID, Group_Assignment, POST_Overall_SEFF.Mean) %>%  
  filter(Group_Assignment == "Intervention") %>% 
  rstatix::t_test(POST_Overall_SEFF.Mean ~ 1, mu = 5.47) 
```
- Post Self-Efficacy Sample is NOT significantly different from the normed (race group) or control. 
- Indication that the intervention may have increase knowledge, but requires longer term work to demonstrate changes in efficacy, and may not be able to adequately account for **starting differences** (paired test)
- Especially given potential differences in Autism severity, and other factors noted as significantly different on the EIPSES paper. 


# 4. Plots

## 4.0 Pre, Post and Control Boxplots
```{r}
EFF.pre_boxplot <- 
  EFF.scored %>% 
  #select(StudyID, Group_Assignment, contains("Total"), contains("Mean")) %>% 
  ggplot(aes(x = Group_Assignment, y = PRE_Overall_SEFF.Mean))+
  geom_boxplot()+
  geom_jitter()

EFF.post_boxplot <- 
  EFF.scored %>% 
  #select(StudyID, Group_Assignment, contains("Total"), contains("Mean")) %>% 
  ggplot(aes(x = Group_Assignment, y = POST_Overall_SEFF.Mean))+
  geom_boxplot()+
  geom_jitter()

library(patchwork)

EFF.pre_boxplot + EFF.post_boxplot + plot_annotation(title = "EFF - Pre/Post Self-Efficacy Scores by Group Assignment")

```


## 4.1 Change in Total Scores (Arrow Plot)
```{r}
EFF.scored %>% 
  arrange(PRE_Overall_SEFF.Mean) %>% 
  mutate(
    Rank_Order = seq_along(PRE_Overall_SEFF.Mean),
    Pct_Order = percent_rank(PRE_Overall_SEFF.Mean)
  ) %>% 
  ggplot(aes(x = Rank_Order)) +
  # 1. Add the arrows (vertical segments)
  geom_segment(aes(
    xend = Rank_Order, 
    y = PRE_Overall_SEFF.Mean, 
    yend = POST_Overall_SEFF.Mean
  ), 
  arrow = arrow(length = unit(.35, "cm")), # Adds the arrow head
  color = "gray", 
  alpha = 0.5) +
  
  # 2. Add the PRE points
  geom_point(aes(y = PRE_Overall_SEFF.Mean, color = "PRE_Overall_SEFF.Mean")) +
  
  # 3. Add the POST points
  geom_point(aes(y = POST_Overall_SEFF.Mean, color = "POST_Overall_SEFF.Mean")) +
  
  # Formatting
  scale_color_manual(values = c("PRE_Overall_SEFF.Mean" = "red", "POST_Overall_SEFF.Mean" = "blue")) +
  labs(
    title = "Individual Self-Efficacy Score Change (Ranked by PRE Score)",
    x = "Participant Rank (Low to High PRE Self-Efficacy)",
    y = "Total Self-Efficacy Score",
    color = "Timepoint"
  ) +
  theme_minimal()
```

### Interpretation
- This chart is interesting to me; because there seems to be an inverted effect, after sorting by rank.
- Participants that **STARTED** at low efficacy generally increased. 
- Participants that **STARTED** at high efficacy, generally decreased.
- Inverse effect of gaining more knowledge?
- Could this be followed up on, by examining **qualitative interview** feedback
  - Suggests further interactive effects, maybe explorable via structural equation modeling. 
