---
title: "SMITH-STUDY: Stress Scales - Pre-Post Analysis"
date: "`r Sys.Date()`"
output: 
  html_document:
    theme: cosmo
    toc: true
    toc_float: true
    code_folding: show
always_allow_html: true

---


# 1. Setup & Data Ingestion
## 1.1 Libraries

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
library(tidyverse)
library(dplyr) #tidyverse
library(janitor)
library(readxl)
library(lubridate)
library(ggpubr)      # Publication-ready plots
library(rstatix)     # T-tests and effect sizes
library(broom)       # Tidy statistical outputs
#2 Scale Validation
library(psych)
library(knitr)
library(kableExtra)
```

## 1.2 Load Data
```{r load dat}
# Define file paths
data_path   <- "../_data/stress data pre post.xlsx"
stress_raw <- read_excel(data_path)

(questions <- names(stress_raw))

# get `...#` at the end of each name for data names

stress_named <- stress_raw %>% 
  rename("StudyID" = "Number assigned by study staff",
         "PRE_Stress.Total" = "Pre stress score",
         "POST_Stress.Total" = "Post stress score") %>% 
  rename_with(~paste0("PRE_Q", (as.numeric(str_extract(., "\\d+$"))-1)), .cols = 2:11) %>% #cols numbered in order so subtract previous columns for ACTUAL question number
  rename_with(~paste0("POST_Q", 
                      (as.numeric(str_extract(., "\\d+$"))-12)) #numbered in order, but SAME questions PRE and POST (subtract previous number of columns (starts on 13-12)
              , .cols = 13:22)
names(stress_named)
```

```{r}
#check all unique strings in questions
uniques <- stress_named %>% select(contains("_Q")) %>% 
  summarise(across(everything(), ~list(unique(.)))) %>% 
  pivot_longer(everything(), names_to = "Question", values_to = "Unique_Responses") %>% 
  unnest(cols = c(Unique_Responses))


unique(uniques$Unique_Responses)
```

## 1.3 Item Scoring and Reverse Coding
```{r}

stress_mutated <- stress_named %>% 
  # 1. Convert text responses to numeric scores
  mutate(across(contains("_Q"), ~case_when(
    . == "Never" ~ 0,
    . == "Almost Never" ~ 1,
    . == "Sometimes" ~ 2,
    . == "Fairly Often" ~ 3,
    . == "Very Often" ~ 4,
    TRUE ~ NA_real_
  ))) %>% 
  # 2. Reverse code specific items
  mutate(across(ends_with(c("Q4", "Q5", "Q7", "Q8")), ~case_when(
    . == 0 ~ 4,
    . == 1 ~ 3,
    . == 2 ~ 2,
    . == 3 ~ 1,
    . == 4 ~ 0,
    TRUE ~ NA_real_
  ))) %>% 
  #3. Participant ID as character
  mutate(StudyID = paste0("stressID_", as.character(StudyID)))

stress_mutated
```

## 1.4 Calculate Survey Totals & Means
```{r}
stress_final <- stress_mutated %>% 
  rowwise() %>%
  # 3. Calculate total scores
  mutate(PRE_Stress.Total = sum(c_across(starts_with("PRE_Q")), na.rm = FALSE),
         POST_Stress.Total = sum(c_across(starts_with("POST_Q")), na.rm = FALSE),
         PRE_Stress.Mean = mean(c_across(starts_with("PRE_Q")), na.rm = FALSE),
         POST_Stress.Mean = mean(c_across(starts_with("POST_Q")), na.rm = FALSE),
         PRE_Stress.StDev = sd(c_across(starts_with("PRE_Q")), na.rm = FALSE),
         POST_Stress.StDev = sd(c_across(starts_with("POST_Q")), na.rm = FALSE)
         ) %>% 
    ungroup()

stress_final
```

## 1.5 Save Data and Write Files
```{r}
#dir.create("../cache", showWarnings = FALSE)
#dir.create("../_output", showWarnings = FALSE)
save(stress_final, file = "../_output/stress_final.RData")
write_csv(stress_final, "../_output/clean_stress_survey_data.csv")
```

# 2 Scale Validation
## 2.1 Load Data 
```{r}
load("../_output/stress_final.RData") #load the cleaned data from script 1

check_table <- stress_final %>% select(StudyID, ends_with(
  c("Total"#,"Mean"
    )))

check_table %>% head() %>% kableExtra::kable() %>% 
  kableExtra::kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = F)
```
## 2.2 Score Distributions (Histogram/Kernal Dist)
```{r}
library(ggridges)

check_table %>% mutate(Split_Pre_Total = ifelse(PRE_Stress.Total > mean(PRE_Stress.Total), "High", "Low")) %>% pivot_longer(cols = PRE_Stress.Total:POST_Stress.Total, names_to = "Measure", values_to = "Value") %>% ggplot(aes(x=Value, y=Measure,fill=Split_Pre_Total))+
  geom_density_ridges()
```

## 2.3 Item-Level Distribution (Dot Plot)

```{r}
stress_final %>% 
  pivot_longer(cols = contains("_Q"), names_to = "Item", values_to = "Value") %>% 
  mutate(
    Condition = if_else(str_detect(Item, "PRE_"), "Pre-Test", "Post-Test"),
    Condition = factor(Condition, levels = c("Pre-Test", "Post-Test")),
    Q_Num = as.numeric(str_extract(Item, "\\d+")),
    Q_Order = factor(Q_Num, levels = 1:10)
    #Composite = if_else(Q_Num %in% c(1, 2, 9, 14), "Parent Competence", "Outcome Expectations")
  ) %>% 
  filter(!is.na(Value)) %>% 
  ggplot(aes(x = Value, y = fct_rev(Q_Order)))+#, color = Composite)) +
  geom_jitter(position = position_jitter(width = 0.2, height = 0.1), alpha = 0.5) +
  stat_summary(fun = "mean", geom = "point", color = "orange", size = 2, shape = 18) +
  facet_wrap(~Condition) +
  scale_x_continuous(breaks = 0:4) +
  theme_minimal() +
  theme(legend.position = "top") +
  labs(
    title = "Item-Level Distributions (Likert Scale 0 - 4 | Never - Very Often)",
    subtitle = "Black diamond (♦) represents the item mean.",
    y = "Question Number",
    x = "Likert Score"
  )
```


## 2. Test Internal Consistency (Cronbach's Alpha)
- Here we will calculate Cronbach's alpha for each scale to check for internal consistency. 
- A general rule of thumb is that an alpha of > 0.7 is acceptable.

```{r}
pre_alpha <- stress_final %>% select(contains("PRE_Q")) %>% psych::alpha()
post_alpha <- stress_final %>% select(contains("POST_Q")) %>% psych::alpha()

## Dataframe of the Cronbach Alpha Tests and Values
alpha_values <- data.frame(
  Scale = c("PRE", "POST"),
  Cronbach_Alpha = c(pre_alpha$total$raw_alpha,
                     post_alpha$total$raw_alpha)
)

alpha_values %>% kableExtra::kable(digits=3,
                                   caption = " Cronbach Alpha Values") %>%
  kableExtra::kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = F)
```

#3. Descriptive Statistics

## 3.1 Descriptiive Table
```{r}
stress_final %>%
  summarise(N = n(),
            Mean_PRE = mean(PRE_Stress.Total, na.rm = TRUE),
            SD_PRE = sd(PRE_Stress.Total, na.rm = TRUE),
            Mean_POST = mean(POST_Stress.Total, na.rm = TRUE),
            SD_POST = sd(POST_Stress.Total, na.rm = TRUE)
            ) %>%
  kable(digits = 2,
        caption = "Summary Table - Group Means and St.Deviations") %>%
  kable_styling(full_width = F)
```

## 3.2 Change in Total Scores (Arrow Plot)
```{r}
stress_final %>% 
  arrange(PRE_Stress.Total) %>% 
  mutate(
    Rank_Order = seq_along(PRE_Stress.Total),
    Pct_Order = percent_rank(PRE_Stress.Total)
  ) %>% 
  ggplot(aes(x = Rank_Order)) +
  # 1. Add the arrows (vertical segments)
  geom_segment(aes(
    xend = Rank_Order, 
    y = PRE_Stress.Total, 
    yend = POST_Stress.Total
  ), 
  arrow = arrow(length = unit(.35, "cm")), # Adds the arrow head
  color = "gray", 
  alpha = 0.5) +
  
  # 2. Add the PRE points
  geom_point(aes(y = PRE_Stress.Total, color = "PRE_Stress.Total")) +
  
  # 3. Add the POST points
  geom_point(aes(y = POST_Stress.Total, color = "POST_Stress.Total")) +
  
  # Formatting
  scale_color_manual(values = c("PRE_Stress.Total" = "red", "POST_Stress.Total" = "blue")) +
  labs(
    title = "Individual Stress Score Change (Ranked by PRE Score)",
    x = "Participant Rank (Low to High PRE Stress)",
    y = "Total Stress Score",
    color = "Timepoint"
  ) +
  theme_minimal()
```

```{r}
stress_final
```


## 3.1 Long Form Data
```{r}
# Structure data long for easier descriptive grouping
stress_long <- stress_final %>%
  select(StudyID, ends_with(".Total")) %>%
  pivot_longer(cols = -StudyID, names_to = "Variable", values_to = "Score") %>%
  mutate(
    Condition = if_else(str_detect(Variable, "PRE_"), "Pre-Test", "Post-Test"),
    Condition = factor(Condition, levels = c("Pre-Test", "Post-Test"))
    )
```


## 3.3 Paired T-Tests

```{r}
# Calculate Descriptives
stress_descrip <- stress_long %>% 
  group_by(Condition) %>% 
  summarize(
    N = n(),
    Missing = sum(Score==0),
    Mean = mean(Score, na.rm=TRUE),
    SD = sd(Score, na.rm=TRUE),
    Min = min(Score, na.rm=TRUE),
    Max = max(Score, na.rm=TRUE),
    .groups = "drop"
  )

stress_descrip %>% 
  kable(caption = "Descriptive Statistics for All Scales", digits = 2) %>% 
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = F)
```

# 4. Statistical Tests (Pre vs. Post Comparisons)
## 4.1 Paired T-Tests & Effect Sizes

We compare the Pre-test and Post-test scores for each of the three composites to observe absolute changes in stress.



```{r}

# Run Paired T-Tests
stress_t_test_results <- stress_long %>%
  #group_by(Composite_Type) %>%
  t_test(Score ~ Condition, paired = TRUE, detailed = TRUE) %>%
  adjust_pvalue(method = "bonferroni") %>%
  add_significance("p.adj")

# Calculate Cohen's d for Effect Size
stress_cohens_d_results <- stress_long %>%
  #group_by(Composite_Type) %>%
  cohens_d(Score ~ Condition, paired = TRUE)

# Combine for final table
# stress_t_test_final <- stress_t_test_results %>%
#   left_join(select(stress_cohens_d_results, 
#                    #Composite_Type, 
#                    effsize, magnitude), 
#             #by = "Composite_Type") %>%
#             by = "Condition") %>%
#   select(
#     #Composite_Type, 
#     statistic, df, p, p.adj, p.adj.signif, effsize, magnitude)

stress_t_test_final <- cbind(
  stress_t_test_results %>% select(statistic, df, p, p.adj, p.adj.signif),
  stress_cohens_d_results %>% select(effsize, magnitude)
)

stress_t_test_final %>% 
  kable(caption = "Paired T-Test Results (Pre vs. Post)", digits = 3) %>% 
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = T)
```

```{r}
stress_long %>%
  #group_by(Composite_Type) %>%
  t_test(Score ~ Condition, paired = TRUE, detailed = TRUE) %>%
  adjust_pvalue(method = "bonferroni") %>%
  add_significance("p.adj")

```

