---
always_allow_html: true
output:
  rmarkdown::github_document: default
  always_allow_html: default
  output:
  md_document:
    variant: gfm
    number_sections: true
knit: (function(inputFile, encoding) {
  rmarkdown::render(inputFile, encoding = encoding, output_dir = "../") })
date: "2025-09-18"
title: "1_read_survey_data"
---

```{r setup, include=FALSE}
.base_dir <- "../" # where output should default to
.base_url <- "" # keep as is
.fig_path <- "plot_images/" # where image output should go

knitr::opts_knit$set(base.dir = .base_dir)
knitr::opts_chunk$set(fig.path = .fig_path,
                      cache.path = '../cache/',  #log files go here
                      message=FALSE, warning=FALSE,
                      cache = TRUE) 
```  

# 0. Setup

## 0.1 Libraries/Packages
```{r}
#load packages
#library(readxl) #to read in data
library(dplyr) #for manipulating data, using PIPES %>% for easier readability
library(stringr)
library(tidyr) #pivoting and manipulating data, structuring data for output
library(readxl) #for reading in data)
library(kableExtra)
# #formatting
# #library(tableone) #make APA style tables for publication
# library(jtools) #APA format - theme
# library(ggplot2) #tailored graphs/charts & data viz for exploration
# #library(patchwork) #for combining ggplot2 charts
# #library(moments) #for skewness and kurtosis calculations
# #library(lme4) #used for mixed effects regression - not currently needed
```

# 0.1 Load Data into R
```{r}
data_files <- list.files("../_data")
data_files
```
- How does the data read in, can we use columns or does it need parsing?
```{r}

fpath <- "../_data/S1_Post -  AFTER Learning Activities_January 11, 2026_19.34.csv"


multi_header_read <- function(fpath){
  #questions <- raw_data[1,]#data comes in with QUESTION SUBLABEL/INSTRUCTIONS as FIRST ROW
  
  #CSV  filetype 
  if(grepl("\\.csv$", fpath)){
    raw_data <- readr::read_csv(file = fpath, col_names = TRUE)
    value_data <- readr::read_csv(file = fpath, col_names = names(raw_data), skip = 3)
    return(value_data)
    }
  #EXCEL
  if(grepl("\\.xlsx$", fpath)){
    raw_data <- readxl::read_excel(path = fpath, col_names = TRUE) #read in the raw survey data
    value_data <- readxl::read_excel(path = fpath, col_names = names(raw_data),skip = 2) #read and skip first row
  return(value_data)
  }
}

pulse1 <- multi_header_read("../_data/PULSE Survey_November 3, 2025_18.24.xlsx")

#glimpse(pulse1)
```

```{r functions for names & cleaning}
clean_names <- function(data.table){
  output.table <- data.table %>% rename_all(~gsub(" ", "_", .)) %>% 
  #replace spaces in colnames with underscores
  rename_all(~gsub("[^[:alnum:]_]", "", .)) 
  #remove special characters from column names
  return(output.table)
  #returns table
  }

pulse1 %>% 
  filter(DistributionChannel=="preview") %>% clean_names() %>% 
  head() %>% 
  kable(caption="Example Output with Cleaned Names") %>% 
  kable_styling(full_width = F)
```

# 1. Data Cleaning and Preparation
```{r}
# read files and name them from the filenames
library(dplyr)
RAWinfo <- multi_header_read("../_data/Information Sheet - AI and Language Learning Survey_November 3, 2025_18.20.xlsx") %>% clean_names()
RAWpre <- multi_header_read("../_data/S1_Pre - BEFORE Learning Activities_November 3, 2025_18.23.xlsx") %>% clean_names()
#RAWpost <- multi_header_read("../_data/S1_Post -  AFTER Learning Activities_November 3, 2025_18.24.xlsx") %>% clean_names()
RAWpost <- multi_header_read("../_data/S1_Post -  AFTER Learning Activities_January 11, 2026_19.34.csv") %>% clean_names()
RAWpulse <- multi_header_read("../_data/PULSE Survey_November 3, 2025_18.24.xlsx") %>% clean_names()
```

### functions
```{r}
likert_to_num <- function(x) {
  case_when(
    x == "Strongly disagree" ~ 1,
    x == "Disagree" ~ 2,
    x == "Somewhat disagree" ~ 2, # Mapping somewhat to 2 or 3 depending on scale, assuming 5 point standard
    x == "Neither agree nor disagree" ~ 3,
    x == "Neither disagree nor agree" ~ 3, # Handle variation
    x == "Somewhat agree" ~ 4,
    x == "Agree" ~ 4,
    x == "Strongly agree" ~ 5,
    TRUE ~ NA_real_
  )
}
```


## PRE-TEST
#### Structure of PreTest
| Section | Original Pattern | New Name Example |
| :--- | :--- | :--- |
| **S-FLES (9 Likert)** | `Q1SFLES13_1` ... `Q3SFLES79_3` | `SFLES_1` ... `SFLES_9` |
| **Cultural Context (Part 1)** | `Q1_1` ... `Q1_4` | `Culture_1` ... `Culture_4` |
| **Cultural Context (Part 2)** | `Q2_134` ... `Q2_437` | `Culture_5` ... `Culture_8` |
| **AI Experience (Likert)** | `Q4_1` ... `Q4_5` | `AI_L_1` ... `AI_L_5` |
| **AI Experience (Categorical)** | `Q1`, `Q2_5`, `Q3` | `AI_Cat_1`, `AI_Cat_2`, `AI_Cat_3` |
| **AI Experience (Open Text)** | `Q2_6_TEXT`, `Q3_6_TEXT` | `AI_Text_1`, `AI_Text_2` |

```{r}
#filter
f.pre <- RAWpre %>% filter(Status != "Survey Preview",
                           Finished =='True')
  #renamte, select needed columns

cat("Start:", nrow(RAWpre), "\n",
    "Filtered:", nrow(RAWpre) - nrow(f.pre), "total rows", "\n",
    "Final:", nrow(f.pre))
```

```{r}
#select, mutate, aggregate (scores)
s.pre <- f.pre %>% 
  select(
    Email = Validation_1,
    Passphrase = Validation_2,
    ResponseId,
    Group = GroupSelect,
    #Questions
    Q1SFLES13_1:Q4_5)


## mutate!!!

m.pre <- s.pre %>%
  mutate_at(vars(Q1SFLES13_1:Q2_437), likert_to_num) %>% 
  mutate_at(vars(Q4_1:Q4_5), likert_to_num) %>% 
  mutate(across(ends_with("TEXT"), ~na_if(., "-99"))) %>%  #remove "-99" values in TEXT
  
  mutate_at(vars(Q1SFLES13_1:Q2_437), as.numeric) %>%
  mutate_at(vars(Q4_1:Q4_5), as.numeric)

# 1. Get names of numeric columns
numeric_cols <- m.pre %>% select(where(is.numeric)) %>% names()
text_cols <- m.pre %>% select(where(is.character)) %>% names()
# 2. Filter these names based on pattern matching and apply renaming

df_pre <- m.pre %>%
  # Rename SFLES items (Q1SFLES13_1 -> SFLES_1, etc.)
    #rename_with(~str_replace(., "Q1SFLES13", "SF.A"), contains("SFLES")) %>%
    #rename_with(~str_replace(., "Q2SFLES46", "SF.B"), contains("SFLES")) %>%
    #rename_with(~str_replace(., "Q3SFLES79", "SF.C"), contains("SFLES"))
  rename_with(~paste0("Pre_SFLES_", 1:3), matches("Q1SFLES13")) %>% 
  rename_with(~paste0("Pre_SFLES_", 4:6), matches("Q2SFLES46")) %>% 
  rename_with(~paste0("Pre_SFLES_", 7:9), matches("Q3SFLES79")) %>% 
  # Rename first set of Cultural items (Q1_1 -> Culture_1)
  rename_with(~str_replace(., "Q1_", "Pre_Culture_"), starts_with("Q1_")) %>% 
  # Rename second set of Cultural items (Q2_134 -> Culture_5, etc.)
  rename_with(~paste0("Pre_Culture_", 5:8), .cols = intersect(numeric_cols, names(select(m.pre, matches("Q2_\\d{3}"))))) %>% 
  # Rename AI Experience Items
  rename_with(~paste0("Pre_AI.EXP_", c(1,2,3,4,5,6,"6T",7,"7T",8)), .cols = intersect(text_cols, names(select(m.pre, matches("Q\\d{1}"))))) %>% 
  # Rename AI Likert items (Q4_1 -> AI_L_1)
  rename_with(~str_replace(., "Q4_", "Pre_AI_L_"), starts_with("Q4_")) %>% 
  # calculate scores..
  rowwise() %>%
  mutate(Pre_FLE_Total = sum(c_across(starts_with("Pre_SFLES")), na.rm = FALSE),
         Pre_FLE_Mean = mean(c_across(starts_with("Pre_SFLES")), na.rm = FALSE),
         Pre_CRE_Total = sum(c_across(starts_with("Pre_CRE")), na.rm = FALSE),
         Pre_CRE_Mean = mean(c_across(starts_with("Pre_CRE")), na.rm = FALSE),
         Pre_AI_L_Total = sum(c_across(starts_with("Pre_AI_L")), na.rm = FALSE),
         Pre_AI_L_Mean = mean(c_across(starts_with("Pre_AI_L")), na.rm = FALSE)
         #Pre_AI.EXP_Total = sum(c_across(starts_with("Pre_AI.EXP_") %>% select(where(is.numeric))), na.rm = FALSE),
         #Pre_AI.EXP_Mean = mean(c_across(starts_with("Pre_AI.EXP_") %>% select(where(is.numeric))), na.rm = FALSE),
         ) %>%
  ungroup()

#add to writing list
write_to_list <- list("PRETEST_results" = df_pre)
```

## POST
```{r}
f.post <- RAWpost %>% filter(DistributionChannel != "preview",
                           Progress == 100)

cat("Start:", nrow(RAWpost), "\n",
    "Filtered:", nrow(RAWpost) - nrow(f.post), "total rows", "\n",
    "Final:", nrow(f.post))
#glimpse(f.post)
```

```{r}
#select, mutate, aggregate scores

df_post <-
  f.post %>% 
  select(
    Email = Validation_1,
    Passphrase = Validation_2,
    ResponseId,
    Group = `2ClassGroupmatch`,
    AI_Activity,
    #Questions
    Q1_1:Q5TAMUsing1013_4) %>% 
  mutate_at(vars(Q1_1:Q5TAMUsing1013_4), likert_to_num) %>% 
  mutate_at(vars(Q1_1:Q5TAMUsing1013_4), as.numeric) %>% 
  rename_with(~paste0("Post_SFLES_", 1:3), matches("Q1_")) %>% 
  rename_with(~paste0("Post_SFLES_", 4:6), matches("Q2_")) %>% 
  rename_with(~paste0("Post_SFLES_", 7:9), matches("Q3_")) %>% 
  rename_with(~str_replace(., "Q4TAMEase\\d{2}_", "Post_TAM.Ease_"), starts_with("Q4TAM")) %>% 
  rename_with(~str_replace(., "Q5TAMUseful\\d{2}_", "Post_TAM.Useful_"), starts_with("Q5TAM")) %>% 
  rename_with(~str_replace(., "Q5TAMUsing\\d{4}_", "Post_TAM.Attitude_"), starts_with("Q5TAM")) %>% 
  #calculate scores
  rowwise() %>%
  mutate(
    Post_FLE_Total = sum(c_across(starts_with("Post_SFLES")), na.rm = FALSE),
    Post_FLE_Mean = mean(c_across(starts_with("Post_SFLES")), na.rm = FALSE),
    Post_TAM_EOU_Mean = mean(c_across(starts_with("Post_TAM.Ease_")), na.rm = FALSE),
    Post_TAM_U_Mean = mean(c_across(starts_with("Post_TAM.Useful_")), na.rm = FALSE),
    Post_TAM_ATT_Mean = mean(c_across(starts_with("Post_TAM.Attitude_")), na.rm = FALSE)
  ) %>%
  ungroup()

#add to write list
write_to_list$post_coded <- df_post
```

```{r}
df_post %>% 
  select(starts_with("Post_")) %>% 
  head() %>% 
  kable(caption="Post Test Coded Data Sample") %>% 
  kable_styling(full_width = F)
```



## Pulse
- Pulse Survey used the same **distribution link** each time. 
- Each Participant ID (or Email, or Mask) can be used as a **group**. 
- And a sequence (Session) created for each occurence. 
- Protocol: Pulse is given **three times**.

```{r}

f.pulse <- RAWpulse %>% filter(Status != "Survey Preview",
                           Finished =='True')

cat("Start:", nrow(RAWpulse), "\n",
    "Filtered", nrow(RAWpulse) - nrow(f.pulse),"total rows", "\n",
    "Final:", nrow(f.pulse))
glimpse(f.pulse)
```


### Code Pulse

```{r}
#unique(f.pulse$Activity_Number)

Activity_Recode <- function(x){
  case_when(
    str_detect(string=x, pattern = "Speaking") == TRUE ~ "Speaking_Activity",
    str_detect(string=x, pattern = "Writing") == TRUE ~ "Writing_Activity",
    str_detect(string=x, pattern = "Vocab") == TRUE ~ "Vocabulary_Activity",
    TRUE ~ as.character(x)
  )
}


s.pulse <- f.pulse %>% 
  select(
    Email = Validationemail_1,
    Passphrase = Validationemail_2,
    ResponseId,
    Group_Number = Group_Number,
    QConditionAI,
    Activity_Number,
    Q1CoreAI_1:Q1CoreAI_7,
    Q1Core_1:Q1Core_5,
    AIGraded = AIGradeRatomg,
    AIGrade_TEXT = AIGradeRatomg_1_TEXT
  )
```

```{r}
m.pulse <- s.pulse %>% 
  #select(everything()) %>% 
  mutate(Activity_Type = Activity_Recode(Activity_Number)) %>%
  mutate_at(vars(Q1CoreAI_1:Q1Core_5), likert_to_num) %>% 
  mutate_at(vars(Q1CoreAI_1:Q1Core_5), as.numeric) %>% 
  mutate(across(ends_with("TEXT"), ~na_if(., "-99"))) %>%  #remove "-99" values in TEXT
  mutate(QConditionAI = ifelse(startsWith(QConditionAI, "Yes"), "AI_Group", "Control")) %>%
  rowwise() %>% 
  mutate(FILTER_VAR = ifelse(
    any(!is.na(across(Q1CoreAI_1:Q1Core_5))), "KEEP", "DROP") #if all values in Q1CoreAI_1:Q1Core_5 are NA, make filter variable "ISSUE_DROP_NAS"
         ) %>% ungroup()

m2.pulse <- m.pulse %>% filter(FILTER_VAR=="KEEP")
cat("Rows dropped due to all NAs in core items:", nrow(m.pulse) - nrow(m2.pulse), "\n",
    "Final Rows in Pulse Data:", nrow(m2.pulse))

#Participant filled out GROUP wrong, manual recode of Q1CORE_AI ->Q1Core
m2.pulse[m2.pulse$ResponseId=="R_2pJtouk5uyUauWt", "QConditionAI"] <- "Control"
m2.pulse[m2.pulse$ResponseId=="R_2pJtouk5uyUauWt", c("Q1Core_1", "Q1Core_2","Q1Core_3","Q1Core_4","Q1Core_5")] <- m2.pulse[m2.pulse$ResponseId=="R_2pJtouk5uyUauWt", c("Q1CoreAI_1", "Q1CoreAI_2","Q1CoreAI_3","Q1CoreAI_4","Q1CoreAI_5")]
m2.pulse[m2.pulse$ResponseId=="R_2pJtouk5uyUauWt", c("Q1CoreAI_1", "Q1CoreAI_2","Q1CoreAI_3","Q1CoreAI_4","Q1CoreAI_5","Q1CoreAI_6","Q1CoreAI_7", "AIGraded")] <- NA
```

```{r}
df_pulse <- m2.pulse %>% 
  group_by(Email) %>%
  mutate(Session = seq_along(Email)) %>%
  ungroup() %>% 
  group_by(Session) %>% 
  rowwise() %>%
  mutate(
    Pulse_CoreAI_Total = sum(c_across(starts_with("Q1CoreAI_")), na.rm = FALSE),
    Pulse_CoreAI_Mean = mean(c_across(starts_with("Q1CoreAI_")), na.rm = FALSE),
    Pulse_Core_Total = sum(c_across(starts_with("Q1Core_")), na.rm = FALSE),
    Pulse_Core_Mean = mean(c_across(starts_with("Q1Core_")), na.rm = FALSE)
  ) %>%
  ungroup() %>% 
  filter(Session < 4)
  


## MULTIPLE SESSSIONS DATA!
# 
# emails <- df_pulse %>% filter(Session > 3) %>% select(Email) %>% distinct() %>% unlist()
# 
# probs <- df_pulse %>% filter(Email %in% emails) %>% arrange(QConditionAI, Group_Number)
# View(probs)
write_to_list$pulse_CODED <- df_pulse
```

```{r}
df_pulse %>% filter(QConditionAI=="AI_Group") %>%
  select(starts_with("Pulse_"), Session) %>% 
  head() %>% 
  kable(caption="Pulse Coded Data Sample") %>% 
  kable_styling(full_width = F)
```

```{r}
df_pulse %>% filter(QConditionAI=="Control") %>%
  select(starts_with("Pulse_"), Session) %>% 
  head() %>% 
  kable(caption="Pulse Coded Data Sample") %>% 
  kable_styling(full_width = F)
```


# Joining and COnditions

```{r}
df_main <- full_join(df_pre, df_post, by = "Email", suffix = c("_Pre", "_Post"))

# Consolidate Group variable (Prioritize Pre, then Post)
df_main <- df_main %>%
  mutate(Group = coalesce(Group_Pre, Group_Post)) %>%
  select(-Group_Pre, -Group_Post)

# 2. Add Condition (Infer from Group ID or Pulse data)
# *Logic*: If student reported using AI in Pulse, they are AI group. 
# We need to create a mapping based on the majority report per group to be safe.

condition_map <- df_pulse %>%
  filter(!is.na(Group_Number)) # Pulse didn't have group easily accessible in select? 
  # Actually Pulse has "Group_Number" in raw. Let's grab it if possible, 
  # but reliance on Email join is safer.

# Join Pulse condition info to Main to confirm assignment
pulse_cond_summary <- df_pulse %>%
  left_join(select(df_main, Email, Group), by="Email") %>%
  group_by(Group) %>%
  summarize(
    AI_Count = sum(str_detect(QConditionAI, "AI"), na.rm=TRUE),
    Trad_Count = sum(str_detect(QConditionAI, "Control"), na.rm=TRUE)
  ) %>%
  mutate(Condition = if_else(AI_Count > Trad_Count, "AI", "Traditional"))

pulse_cond_summary %>% 
  kable(caption="Condition Assignment Summary by Group") %>% 
  kable_styling(full_width = F)
```

```{r}
# Apply Condition to Main
df_main <- df_main %>%
  left_join(select(pulse_cond_summary, Group, Condition), by = "Group")

# Anonymize
df_main <- df_main %>%
  mutate(Participant_ID = paste0("P", row_number()))

# create separate df with email and id for joining later
df_main_match_id_email <- df_main %>% select(Email, Participant_ID)


# DROP EMAIL FROM ALL DATA
df_pre <- df_pre %>%
  left_join(df_main_match_id_email, by="Email") %>%
  select(Participant_ID, everything(), -Email)

df_post <- df_post %>%
  left_join(df_main_match_id_email, by="Email") %>%
  select(Participant_ID, everything(), -Email)

#drop email
df_main <- df_main %>% select(Participant_ID, Condition, everything(), -Email)

# Add IDs to Pulse
df_pulse <- df_pulse %>%
  left_join(df_main_match_id_email, by="Email") %>% 
  select(Participant_ID, everything(), -Email) %>%
  filter(!is.na(Participant_ID)) # Remove pulse entries that don't match main list
```

```{r}
#final list, R data and write to xlsx
write_to_list <- list(
  "MAIN_CLEANED" = df_main,
  "PULSE_CLEANED" = df_pulse,
  "PRETEST_CLEANED" = df_pre,
  "POSTTEST_CLEANED" = df_post
)

##
dir.create("../_output/", showWarnings = FALSE)
write_rds(write_to_list, file = "../_output/clean_data.rds")
writexl::write_xlsx(write_to_list, path = "../_output/cleaned_survey_data.xlsx")
```

# -------
# APPENDIX
---------------------------
```{r}
#unique(pre$Q1SFLES13_3)
recode_lik <- function(x){
  case_match(.x = x,
            .default = NA,
            "Strongly agree" ~ "5",
            "Somewhat agree" ~ "4",
            "Neither agree nor disagree" ~ "3",
            "Neither disagree nor agree" ~ 3, # Handle variation
            "Somewhat disagree" ~ "2",
            "Strongly disagree" ~ "1")
}

```