---
always_allow_html: true
output:
  rmarkdown::github_document: default
  always_allow_html: default
  output:
  md_document:
    variant: gfm
knit: (function(inputFile, encoding) {
  rmarkdown::render(inputFile, encoding = encoding, output_dir = "../") })
title: "3. Descriptives & Baseline Equivalence"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

base_dir <- "../" # where output should default to
base_url <- "" # keep as is
fig_path <- "_plot_images/" # where image output should go

knitr::opts_knit$set(base.dir = base_dir)
knitr::opts_chunk$set(fig.path = fig_path,
                      cache.path = '../cache/',  #log files go here
                      message=FALSE, warning=FALSE,
                      cache = TRUE) 
```  

```{r}
library(tidyverse)
library(knitr)
library(kableExtra)
library(rstatix) # Great for quick statistical tests
library(ggpubr)  # For publication-ready plots

load("../_output/stress_final.RData") #load the cleaned data from script 1
```

#1. Participant Descriptives

```{r}
stress_final %>%
  summarise(N = n(),
            Mean_PRE = mean(PRE_Stress.Total, na.rm = TRUE),
            SD_PRE = sd(PRE_Stress.Total, na.rm = TRUE),
            Mean_POST = mean(POST_Stress.Total, na.rm = TRUE),
            SD_POST = sd(POST_Stress.Total, na.rm = TRUE)
            ) %>%
  kable(digits = 2,
        caption = "Summary Table") %>%
  kable_styling(full_width = F)
```



# 2. Run Cronbach’s alpha for internal consistency
- Here we will calculate Cronbach's alpha for each scale to check for internal consistency. 
- A general rule of thumb is that an alpha of > 0.7 is acceptable.

```{r}
pre_alpha <- stress_final %>% select(contains("PRE_Q")) %>% psych::alpha()
post_alpha <- stress_final %>% select(contains("POST_Q")) %>% psych::alpha()

## Dataframe of the Cronbach Alpha Tests and Values
alpha_values <- data.frame(
  Scale = c("PRE", "POST"),
  Cronbach_Alpha = c(pre_alpha$total$raw_alpha,
                     post_alpha$total$raw_alpha)
)

alpha_values %>% kableExtra::kable(digits=3,
                                   caption = " Cronbach Alpha Values") %>%
  kableExtra::kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = F)
```

#3. Descriptive Statistics & Paired T-Tests

```{r}
# Basic Paired T-test
t_test_results <- t.test(
  stress_final$POST_Stress.Total, 
  stress_final$PRE_Stress.Total, 
  paired = TRUE
)

t_test_results %>% tidy() %>% 
kable(digits =3, caption = "Paired T-Test Results") %>% 
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = F)
```
- **Result** is non-significant, p = 0.57

### Pivot Data for RM ANOVA
- Will end with same result, but just in case you add a Timepoint 3 in the future
```{r}
#pivot_longer stress, split on PRE and POST into condition groups 
stress_long <- stress_final %>%
  pivot_longer(cols = -StudyID,
               names_to = c("Timepoint", "Question"), 
               values_to = "Value",
              names_sep = "_") %>% 
  mutate(Timepoint = factor(Timepoint, levels = c("PRE", "POST")))
# View a sample of the long data
stress_long %>% glimpse()
```

```{r}
## T-TESTS From LONG FRAME

# t_test_results <- stress_long %>% 
#   filter(grepl(pattern = "Total", Question)) %>%
#   t_test(Value ~ Timepoint, paired = TRUE, detailed = TRUE) %>%
#   adjust_pvalue(method = "bonferroni") %>%
#   add_significance("p.adj")
# 
# # Calculate Cohen's d for Effect Size
# cohens_d_results <- stress_long %>% 
#   filter(grepl(pattern = "Total", Question)) %>%
#   cohens_d(Value ~ Timepoint, paired = TRUE)
# 
# 
# t_test_final <- merge.data.frame(t_test_results, cohens_d_results) %>%
#   select(Composite_Type, statistic, df, p, p.adj, p.adj.signif, effsize, magnitude)
```


## RM ANOVA
```{r}
library(ez)
anova_results <- stress_long %>% filter(Question == "Stress.Total") %>% 
  ezANOVA(
  dv = Value,
  wid = StudyID,
  within = Timepoint,
  detailed = TRUE
)

print(anova_results)
```
- **Result** is non-significant, p = 0.57


## Plots 

```{r}
stress_long %>% filter(grepl(x = Question, pattern = "Q")) %>%
ggplot(aes(x = Question, y=Value, color = Timepoint, fill = Timepoint)) +
  geom_jitter(alpha = 0.6, width = 0.5, height = 0)+
  geom_boxplot(width = 0.3, outlier.shape = NA, alpha = 0.5) +
  #geom_line(aes(group = StudyID), color = "gray", alpha = 0.3) + # Lines connecting individuals
  #geom_point(aes(color = Timepoint), alpha = 0.6) +
  theme_minimal() +
    theme(legend.position = "bottom") +
  labs(
    title = "Change in Stress Scores: PRE vs. POST",
    x = "Total Stress Score",
    y = "Assessment Time"
  )
```

```{r}
stress_long %>% filter(grepl(x = Question, pattern = "Total")) %>%
ggplot(aes(x = Timepoint, y=Value, color = Timepoint, fill = Timepoint)) +
  geom_point(alpha = 0.6, width = 0.1, height = 0)+
  geom_boxplot(width = 0.3, outlier.shape = NA, alpha = 0.5) +
  #geom_line(aes(group = StudyID), color = "gray", alpha = 0.3) + # Lines connecting individuals
  #geom_point(aes(color = Timepoint), alpha = 0.6) +
  theme_minimal() +
    theme(legend.position = "right") +
  labs(
    title = "Change in Stress Scores: PRE vs. POST",
    x = "Total Stress Score",
    y = "Assessment Time"
  )
```
